The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-12-23 doc/SA-Hermaphrodites-wDemography-R1.tex
minor edits

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying parameter conditions under which SA polymorphism is maintained {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, highly unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{there is negative correlation between different fitness components} \citep{CharlesworthHughes2000, ConnallonHall2018}.

\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection and gene flow} have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On the one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter), \textcolor{blue}{is an important form of conflicting selection that arises when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness, indicating SA fitness effects for multilocus genotypes (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}). Moreover, polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

@@

\textcolor{blue}{While significant efforts have been made to understand when balanced SA polymophisms can be maintained and to detect SA genetic variation from empirical data, much less consideration has been given to the demographic consequences of the maladaptation caused by SA selection}. Most population genetic models, including models of SA selection, model the relative fitness of genotypes in populations of constant (and often infinite) size, and do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{that link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). Over the past 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. In particular, theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic changes accompanying allele frequency changes due to selection. Extensions of these theories to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects} all highlight the sometimes counterintuitive demographic consequences of adaptive evolution that are missed by classical population genetics models (see \citealt{fussmann2007eco} and \citealt{govaert2019eco} for comprehensive reviews). More generally, eco-evolutionary approaches contribute a }

@@

The population state at time $t$ is described by a population state vector, $\tilde{\mbf{n}}(t)$, which is ordered by how individuals were produced (selfing vs.~outcrossing), then by genotype, and finally by stage. For a single diallelic locus with alleles $A$ and $a$, we have three genotypes ($AA,\, Aa,\, aa$; $g = 3$), giving the population state vector is:
\begin{linenomath*}
\begin{equation} \label{eq:PopStateVec}
	\tilde{\mbf{n}}(t) =  \left[
								\begin{array}{c}
									\mbf{n}^{S}_{AA}(t) \\
									\mbf{n}^{S}_{Aa}(t) \\
									\mbf{n}^{S}_{aa}(t) \\ \hline
									\mbf{n}^{X}_{AA}(t) \\
									\mbf{n}^{X}_{Aa}(t) \\
									\mbf{n}^{X}_{aa}(t) \\ 
						\end{array} \right],
\end{equation}
\end{linenomath*}

@@

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contribution of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime}_{i}$, respectively. Deviations from this assumption are straightforward to incorporate but beyond the scope of this paper.

\textcolor{blue}{If the survival or fertility rates are a function of the population density, $\tilde{\mbf{n}}(t)$, then the model will be a density dependent model and the population projection matrix becomes a function of the full population vector, $\tilde{\mbf{A}}[\tilde{\mbf{n}}(t)]$. Below, we limit our analyses to consider density-independent survival and fertility rates. The population projection matrix is then only a function of the proportional population vector, $\tilde{\mbf{A}}[\tilde{\mbf{p}}(t)]$, and the model is referred to as a frequency-dependent model. In \ref{App:DensDep} we construct and analyze a density-dependent version of the model and show that with stage-independent density-dependent growth, our main conclusions remain intact.}

@@

\textcolor{red}{Lotte, I've reworded this to avoid dioecious language (males/females) and to emphasize that what really matters is that outcross matings are random, and selfing rates are genotype-independent... please check to make sure you agree with my edits} \textcolor{blue}{ For the purpose of this article, we assume that all outcross matings are random. The model can be extended to consider various forms of assortative mating beyond genotype-dependent selfing (see Online Supplement). In addition, we assume that individuals produced by selfing or by outcrossing do not differ in their mating success, though they can differ in their surival (see below). Given these assumptions, the allele frequencies in the male mating population can obtained by simply summing the vectors of individuals produced through selfing and outcrossing:
\begin{linenomath*}
\begin{equation}
 	\mbf{p}^{\prime}=\frac{(\mbf{n}^X + \mbf{n}^S)}{\| (\mbf{n}^X + \mbf{n}^S)\|}.
 \end{equation}
\end{linenomath*}}

\textcolor{blue}{Individuals contribute  gametes to the population outcrossing gamete pool according to their fertility, which depends on their stage and genotype at the SA locus. The allele frequencies in the male gamete pool are then calculated as follows (see Online Supplementary Materials for a derivation of this equation),
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) \propto 
				\left(
		\begin{array}{c}
			\mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{AA}\mbf{p}^{\prime}_{AA}+ \frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa}  \\
		\frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa} + \mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{aa}\mbf{p}^{\prime}_{aa}\\
		\end{array} \right),
\end{equation}
\end{linenomath*}
where equality is obtained by normalizing the frequencies such that they add to one. The first entry sums the the stage-specific contributions to the gamete pool of heterozygotes and $AA$ homozygotes; the second entry sums the stage-specific contributions to the gamete pool of heterozygotes and $aa$ homozygotes. }

\textcolor{blue}{ We now have all the ingredients to calculate the allele frequencies in the male gamete pool for the two-stage model in Eq(\ref{eq:F_male}), Eq(\ref{eq:maleGametePool}), and Table \ref{tab:Fitness}. Denoting the proportion of adults of genotype $i$ with $p^A_i$, the allele frequencies in the male gamete pool are given by 
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool_example}
\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
				\frac{1}{(1-s_m)p^A_{aa}+ (1-hs_m)p^A_{Aa}+p_{aa}}\left(
		\begin{array}{c}
(1-s_m)p^A_{AA}	+\frac{1}{2}(1-hs_m)p^A_{Aa}\\
p^A_{aa}+\frac{1}{2}(1-hs_m)p^A_{Aa}
		\end{array} \right).
\end{equation}
\end{linenomath*}
Note that this equation is identical to the classical population genetic results for frequencies of male gametes in a sexually antagonistic model, with the caveat that only reproductively active adults contribute to the gamete pool. Eq(\ref{eq:maleGametePool_example}) increasingly deviates from the classical population genetics result if additional reproductive adult age or stage classes are included (see Eq(\ref{eq:maleGametePool})).  }

@@

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. \textcolor{blue}{Although we use density-independent survival and fertility rates, the allele frequency dynamics introduce frequency-dependence into the model. As a consequence, }the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see chapter 17 in \citealt{Caswell2001}). The intrinsic population growth rate after convergence, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$. We note, however, that if the ecological component of the model is non-linear, more exotic dynamics are possible \citep{de2020matrix}.

@@

%%%%%%%%%%%%%%%%%%%%%
% Figure 2
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{extinctionThresholdsFig}
 \caption{\footnotesize{Illustration of parameter space for SA polymorphism and extinction thresholds predicted by the model. Balanced SA polymorphisms can be maintained in the funnel-shaped region between the invasion conditions for each SA allele (dark solid lines). However, for some parameter conditions, populations will ultimately go extinct (red shaded regions) due to reduced female fitness resulting from the male-beneficial/female-deleterious allele that is either segregating as a balanced polymorphism (inside the funnel), or becomes fixed (area below the funnel). "Demographically viable polymorphic parameter space" corresponds to the area inside the funnel that is also to the left of the extinction threshold for a given fertility value. Results are shown for three different population selfing rates ($C = \{0,\,1/4,\,1/2\}$), and two dominance scenarios (additivity, where $h = 1/2$, and dominance reversal, where $h = 1/4$); extinction thresholds are illustrated for three different values of female fecundity ($f$ values annotated on each panel). \textcolor{blue}{Predicted population intrinsic growth rates ($\lambda$) for the same parameter conditions are presented in fig.~S1, in the Online Supplementary Material.}}}
 \label{fig:extThresholds}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%

@@

However, a key prediction from the evolutionary demographic model is that large fractions of SA polymorphic parameter space can be demographically inviable (fig.~\ref{fig:extThresholds}). The location of the extinction threshold, where the population intrinsic growth rate $\lambda = 1$, is primarily determined by the fertility parameter ($f$) but is also influenced by the population selfing rate ($C$) and dominance of the SA alleles ($h$).

@@

Interestingly, a polymorphic chromosomal inversion (inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. inv6 segregates at moderate frequency (about $8\%$), and carriers suffer an approximately $30 \%$ loss in pollen viability, but also increased flower (and therefore ovule and pollen) production. The genetic basis of these effects remain unclear -- though they likely polygenic -- but the net result is a "supergene" with remarkably strong effects on both female and male fertility that segregates as a single diallelic locus \textcolor{blue}{(i.e., with two alleles, wild-type and inverted)}. We estimated selection coefficients for the effect of inv6 on pollen production (i.e., taking into account the simultaneous effect on flower number) and ovule production from the data reported in \citet{LeeKelly2015} under the relatively conservative assumptions that pollen/ovule production is proportional to flower number, and  additive SA fitness effects (dominance coefficients could not be estimated from field data; see \ref{App:Mimulus} for details of estimating selection coefficients). Under these assumptions, the average selection coefficient for inv6 on flower production across two years in the field was $(s_f,s_m) = (0.3,0.31)$. As a final proof-of-concept for our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, inv6 appears to fall in demographically viable SA polymorphic parameter space (see fig.~\ref{fig:mimulusFig}), as might be expected given its observed frequency in the Iron Mountain population.

@@

@@

The first and central finding of our study is that when population intrinsic growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. Since intrinsic growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exception of recently introduced, invasive populations), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. In addition, we find that much of the parameter space where a male-beneficial allele \textcolor{blue}{fixes} is demographically inviable for populations whose intrinsic growth rates are close to one prior to invasion. This demographic consequence of masculinization is rarely considered \citep{hitchcock2020gene}.  Moreover, these findings complement recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be high \citep{BergerEtAl2014, Connallon2015}.

@@

Despite the demographic pitfalls associated with SA alleles, our case study using {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space. The example also appears to provide some empirical support for the conjecure that locally-adapted populations are more likely to harbor SA polymorphisms than marginal or non-locally adapted ones; inviable polymorphic parameter space only occured when using the demographic data for non-local low-elevation perennial populations. Although we cannot make concrete predictions for inv6 in the Iron Mountain population in which it was observed, it is interesting that the estimated SA fitness effects place this polymophic inversion squarely in demographically viable polymorphic parameter space predicted by our model. The available data do not allow for confident estimation of selection coefficients for inv6 (and even require making assumptions about dominance), yet our theoretical predictions are encouragingly consistent with the available data that inv6 is segregating at intermediate frequencies in the large and locally adapted Iron Mountain population.

>>>>

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying parameter conditions under which SA polymorphism is maintained {\itshape and} the population growth rate remains positive. \textcolor{blue}{A sexually antagonistic polymorphism is a polymorphism of two alleles with antagonistic effects on the two sexes, that is, one of the two coexisting allele confers a fitness benefit to females and reduces male fitness, and the other allele confers a fitness benefit to males and reduces female fitness. } Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{there is a negative correlation between different fitness components} \citep{CharlesworthHughes2000, ConnallonHall2018}.

\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection and gene flow between different classes of individuals} have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On the one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that arises when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness, indicating SA fitness effects for multilocus genotypes (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}). Moreover, polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

@@

\textcolor{blue}{While significant efforts have been made to understand when balanced SA polymophisms can be maintained and to detect SA genetic variation from empirical data, much less consideration has been given to the demographic consequences of the maladaptation caused by SA selection}. Most population genetic models, including models of SA selection, model the relative fitness of genotypes in populations of constant (and often infinite) size, and do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{that link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). Over the past 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. In particular, theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic changes accompanying allele frequency changes due to selection. Extensions of these theories to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects} all highlight the sometimes counterintuitive demographic consequences of adaptive evolution that are missed by classical population genetics models. }

@@

The population state at time $t$ is described by a population state vector, $\tilde{\mbf{n}}(t)$, which is ordered by how individuals were produced (selfing vs.~outcrossing), then by genotype, and finally by stage. For a single diallelic locus with alleles $A$ and $a$, we have three genotypes ($AA,\, Aa,\, aa$; $g = 3$), giving the population state vector:
\begin{linenomath*}
\begin{equation} \label{eq:PopStateVec}
	\tilde{\mbf{n}}(t) =  \left[
								\begin{array}{c}
									\mbf{n}^{S}_{AA}(t) \\
									\mbf{n}^{S}_{Aa}(t) \\
									\mbf{n}^{S}_{aa}(t) \\ \hline
									\mbf{n}^{X}_{AA}(t) \\
									\mbf{n}^{X}_{Aa}(t) \\
									\mbf{n}^{X}_{aa}(t) \\ 
						\end{array} \right],
\end{equation}
\end{linenomath*}

@@

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contributions of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime}_{i}$, respectively. Deviations from this assumption are straightforward to incorporate but beyond the scope of this paper.

\textcolor{blue}{If the survival or fertility rates are a function of the population density, $\tilde{\mbf{n}}(t)$, then the model will be a density dependent model and the population projection matrix becomes a function of the full population vector, $\tilde{\mbf{A}}[\tilde{\mbf{n}}(t)]$. Below, we limit our analyses to consider density-independent survival and fertility rates. The population projection matrix is then only a function of the proportional population vector, $\tilde{\mbf{A}}[\tilde{\mbf{p}}(t)]$, and the model is referred to as a frequency-dependent model. In \ref{App:DensDep} we construct and analyze a density-dependent version of the model and show that with density-dependent survival affecting all stages equally, our main conclusions remain intact.}

@@

 \textcolor{blue}{ For the purpose of this article, we assume that all outcross matings are random. In addition, we assume that individuals produced by selfing or by outcrossing do not differ in their mating success, though they can differ in their surival (see above). Given these assumptions, the allele frequencies in the male mating population can obtained by simply summing the vectors of individuals produced through selfing and outcrossing:
\begin{linenomath*}
\begin{equation}
 	\mbf{p}^{\prime}=\frac{(\mbf{n}^X + \mbf{n}^S)}{\| (\mbf{n}^X + \mbf{n}^S)\|}.
 \end{equation}
\end{linenomath*}}

\textcolor{blue}{Individuals contribute  gametes to the population outcrossing gamete pool according to their fertility, which depends on their stage and genotype at the SA locus. The allele frequencies in the male gamete pool are then calculated as follows (see Online Supplementary Materials for a derivation of this equation),
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) \propto 
				\left(
		\begin{array}{c}
			\mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{AA}\mbf{p}^{\prime}_{AA}+ \frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa}  \\
		\frac{1}{2} \mbf{1}^{\intercal}_{\omega}  \mbf{F}^{\prime}_{Aa} \mbf{p}^{\prime}_{Aa} + \mbf{1}^{\intercal}_{\omega}\mbf{F}^{\prime}_{aa}\mbf{p}^{\prime}_{aa}\\
		\end{array} \right),
\end{equation}
\end{linenomath*}
where equality is obtained by normalizing the frequencies such that they add to one. The first entry sums the  stage-specific contributions to the gamete pool of heterozygotes and $AA$ homozygotes; the second entry sums the stage-specific contributions to the gamete pool of heterozygotes and $aa$ homozygotes. }

\textcolor{blue}{ We now have all the ingredients to calculate the allele frequencies in the male gamete pool for the two-stage model in Eq(\ref{eq:F_male}), Eq(\ref{eq:maleGametePool}), and Table \ref{tab:Fitness}. Denoting the proportion of reproducing adults of genotype $i$ with $p^R_i$, the allele frequencies in the male gamete pool are given by 
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool_example}
\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
				\frac{1}{(1-s_m)p^R_{aa}+ (1-hs_m)p^R_{Aa}+p_{aa}}\left(
		\begin{array}{c}
(1-s_m)p^R_{AA}	+\frac{1}{2}(1-hs_m)p^R_{Aa}\\
p^R_{aa}+\frac{1}{2}(1-hs_m)p^R_{Aa}
		\end{array} \right).
\end{equation}
\end{linenomath*}
Note that this equation is identical to the classical population genetic results for frequencies of male gametes in a sexually antagonistic model, with the caveat that only reproductively active adults contribute to the gamete pool. Eq(\ref{eq:maleGametePool_example}) increasingly deviates from the classical population genetics result if additional reproductive adult age or stage classes are included (see Eq(\ref{eq:maleGametePool})).  }

@@

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. \textcolor{blue}{Although we use density-independent survival and fertility rates, the allele frequency dynamics introduce frequency-dependence into the model. As a consequence, }the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see chapter 17 in \citealt{Caswell2001}). The intrinsic population growth rate after convergence, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$.

@@

%%%%%%%%%%%%%%%%%%%%%
% Figure 2
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{extinctionThresholdsFig}
 \caption{\footnotesize{Illustration of parameter space for SA polymorphism and extinction thresholds predicted by the model. Balanced SA polymorphisms can be maintained in the funnel-shaped region between the invasion conditions for each SA allele (dark solid lines). However, for some parameter conditions, populations will ultimately go extinct (red shaded regions) due to reduced female fitness resulting from the male-beneficial/female-deleterious allele that is either segregating as a balanced polymorphism (inside the funnel), or becomes fixed (area below the funnel). "Demographically viable polymorphic parameter space" corresponds to the area inside the funnel that is also to the left of the extinction threshold for a given fertility value. Results are shown for three different population selfing rates ($C = \{0,\,1/4,\,1/2\}$), and two dominance scenarios (additivity, where $h = 1/2$, and dominance reversal, where $h = 1/4$). Extinction thresholds are illustrated for three different values of female fecundity ($f$ values annotated on each panel); \textcolor{blue}{ the coexistence boundaries are unaffected by the change in $f$.} \textcolor{blue}{Surface plots of predicted population intrinsic growth rates ($\lambda$) for the same parameter conditions are presented in fig.~S1, in the Online Supplementary Material.}}}
 \label{fig:extThresholds}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%

@@

However, a key prediction from the evolutionary demographic model is that large fractions of SA polymorphic parameter space can be demographically inviable (fig.~\ref{fig:extThresholds}). The location of the extinction threshold, where the population intrinsic growth rate $\lambda = 1$, is primarily determined by the fertility parameter ($f$) but is also influenced by the population selfing rate ($C$), and dominance of the SA alleles ($h$).

@@

Interestingly, a polymorphic chromosomal inversion (inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. inv6 segregates at moderate frequency (about $8\%$), and carriers suffer an approximately $30 \%$ loss in pollen viability, but also increased flower (and therefore ovule and pollen) production. The genetic basis of these effects remain unclear -- though they are likely polygenic -- but the net result is a "supergene" with remarkably strong effects on both female and male fertility that segregates as a single diallelic locus \textcolor{blue}{(i.e., with two alleles, wild-type and inverted)}. We estimated selection coefficients for the effect of inv6 on pollen production (i.e., taking into account the simultaneous effect on flower number) and ovule production from the data reported in \citet{LeeKelly2015} under the relatively conservative assumptions that pollen/ovule production is proportional to flower number, and  additive SA fitness effects (dominance coefficients could not be estimated from field data; see \ref{App:Mimulus} for details of estimating selection coefficients). Under these assumptions, the average selection coefficient for inv6 on flower production across two years in the field was $(s_f,s_m) = (0.3,0.31)$. As a final proof-of-concept for our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, inv6 appears to fall in demographically viable SA polymorphic parameter space (see fig.~\ref{fig:mimulusFig}), as might be expected given its observed frequency in the Iron Mountain population.

@@

\textcolor{blue}{In Appendix B we recreate Figure \ref{fig:mimulusFig} for a density-dependent model, replacing the heatmap of population growth rate with a heatmap of equilibrium population densities. We find that for a density-dependent term that affects survival of all (st)age classes in the same way, the coexistence boundaries are unchanged and the heatmap has the same shape. }

@@

The first and central finding of our study is that when population intrinsic growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. Since intrinsic growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exception of recently introduced, invasive populations), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. In addition, we find that much of the parameter space where a male-beneficial allele \textcolor{blue}{fixes} is demographically inviable for populations whose intrinsic growth rates are close to one prior to invasion. This demographic consequence of masculinization is rarely considered \citep{hitchcock2020gene}.  Moreover, these findings complement recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be higher \citep{BergerEtAl2014, Connallon2015}.

@@

Despite the demographic pitfalls associated with SA alleles, our case study using {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space. The example also appears to provide some empirical support for the conjecture that locally-adapted populations are more likely to harbor SA polymorphisms than marginal or non-locally adapted ones; inviable polymorphic parameter space only occured when using the demographic data for non-local low-elevation perennial populations. Although we cannot make concrete predictions for inv6 in the Iron Mountain population in which it was observed, it is interesting that the estimated SA fitness effects place this polymophic inversion squarely in demographically viable polymorphic parameter space predicted by our model. The available data do not allow for confident estimation of selection coefficients for inv6 (and even require making assumptions about dominance), yet our theoretical predictions are encouragingly consistent with the available data that inv6 is segregating at intermediate frequencies in the large and locally adapted Iron Mountain population.
